Interactive Visualization of Hierarchically Structured Data

### 2016-12-06 treelapse.Rmd
Revisions to case studies and conclusion

---
title: "Interactive Visualization of Hierarchically Structured Data"
author: "Kris Sankaran"
date: "`r Sys.Date()`"
output:
     bookdown::pdf_book:
bibliography: ["bibliography.bib"]
link-citations: yes
---

@@

In the DOI sankey, we split each edge in the DOI Tree across different
groups. For example, suppose we have the average counts for treatment
and control groups at each tree node. Every edge in the tree is split
into two colors^[We use the colorbrewer palette to facilitate
readability], with relative widths of the different colors reflecting
differences in sizes for the two groups. The overall width of each
edge represents the sum of values across all groups.

@@

The final display currently implemented in the package is the natural
converse of the Timebox Trees display. Rather than defining visual
queries in terms of time series, it defines queries using nodes in the
tree. For this reason, we call the display Treeboxes. Rather than
focusing on the intersection of brushes, as in Timebox Trees, we focus
on the union of brushed over nodes. This allows us to highlight series
associated with nodes on distant subtrees. This display is also suited
for the microbial dynamics problem. For example, by highlighting all
nodes at one taxonomic level in the tree, we can easily summarize the
time series pattern for all the taxa at that level. Alternatively,
focusing on all the children below a single node makes it possible to
see how much correlation and competition there is between
taxonomically similar microbes. As in the Timebox Trees display, a
search box and pan-zoom scented widget are provided.

@@

We now delve into concrete examples of applying treelapse to real
data. Our goals are is to illustrate potential workflows that
incorporate treelapse, describe the formulation of questions that can
be naturally investigated with our methods, and provide example
interpretations for specific treelapse output. Our examples are also
chosen to reflect the range of problem domains to which the package
can be applied -- though it was motivated by applications to the
microbiome, it is not tied to it. More importantly, we argue that the
visualization principles described earlier can substantively improve
the practice of data analysis in the class of problems we have focused
attention on.

@@

[@dethlefsen2008pervasive] investigates the effect of antibiotics on microbial
community composition from an ecological perspective. The study tracks the
microbiome of *x* patients across *y* weeks (todo: give the numbers), with
antibiotic two time courses introduced during * and *, respectively. Discerning
the variation in resilience across microbes is important, considering the
community's increasing acceptance of the idea that microbes are central to human
health and not just disease.

We approach the data using the linked time and treebox views, after first
filtering low variance taxa and taking an $asinh$ transformation. An initial
view, Figure \@ref(fig:antibioticoverview), reveals two dramatic drops in the
overall microbial abundance time series at the time of the antibiotics time
courses. Two more subtle effects are also suggested from this view,

@@

Next, using the scented widget, we focus on the window around the
second antibiotic treatment. We apply the treebox display to compare
then behavior of different families of Firmicutes, Lachnospiraceae and
Ruminoccocus. We suspect these taxa are associated with the delayed
recovery after the second time course. To inspect this, we these
family names in the search box to isolate their positions on the tree,
and then we apply brushes to highlight the series that contribute to
these higher-level families. The resulting view is given by Figure
\@ref(fig:antibioticfirmicutes)

@@

Alternatively, we can summarize each node by the average of its descendants --
this brings attention to individual microbes that may be underlying some of the
broader taxonomic patterns we have noted when studying the subtree sums. For
example, in Figure \@ref(fig:ruminococcus), we highlight all families below the
Ruminococcus, suggesting that the decrease due to antibiotics occurs uniformly
across almost all families. A point that was not evident in the earlier
sum-across-descendants view is that, after the second treatment of antibiotics,
a few of the Ruminoccocus families recover more rapidly than the rest, for
example the Unc095d3 (highlighted in red) are only briefly affected. In
contrast, most families seem to recover in unison after the first treatment.

@@

Further, note that in this subtree averages view, the tree display has
changed. This is because, at each branching point, we place the node
with larger average value on the left. This places the
Verrucomicrobiae at the far left, which seem to have large abundance
over many time points. This phylum had been previously obfuscated,
because there are not many leaves associated with this phylum, hence
the sum was small. Interestingly, the abundance of these microbes
seems to *increase* after the first antibiotics treatment.  Be
cautious, however, that the average over only a few Verrucomicrobiae
species will be a more variable estimate than the averages over the
more prevalent phyla.

@@

The study [@digiulio2015temporal] tracked the abundance of microbes in the
during vaginal microbiome during pregnancy in an effort to study relationships
between microbial community composition and preterm birth. Ideally, it would be
possible to develop clear microbial signatures associated with preterm births.

Unlike the antibiotics study, we have measurements across more individuals than
we could reasonably inspect one at a time. While we could average across all
individuals, we will take our cue from [@digiulio2015temporal] and place each
sample into one of five Community State Types (CSTs), identified via k-medoids.
Using a linear model, one of these CSTs (CST 4) was found to be more diverse,
further it appeared associated with preterm births. Here, we corroborate this
finding using exploratory views.

Therefore, our focus here is on the differential abundance question,
rather than dynamics.  We would like to provide visual representations
of differential abundance across CSTs and also between preterm and
non-preterm births. [@digiulio2015temporal] interpreted its CSTs using
a heatmap, with microbes ordered according to a hierarchical
clustering.  By using the DOI sankey instead, we can interpret the
CSTs in their taxonomic context and at multiple scales of taxonomic
resolution. Further, while [@digiulio2015temporal] focused on
identifying associations between preterm births and CSTs -- presumably
because testing individual microbes loses power -- we can compare
microbe abundances between preterm and non-preterm samples along
subtrees, without requiring CSTs as an intermediary.

In Figure \@ref(fig:pretermcsts), we compare the 5 CSTs according to
their values along the subtree. Specifically, we took the average of
all samples within each CST to define values at the (microbe-level)
leaf nodes, and then aggregated the counts up to the root. It is
immediately clear that samples from CST 4 have much more taxonomic
diversity. Further, focusing on the Lactobacillaceae family, we note
that the differential abundnace of these microbes distinguishes th
remaining CSTs, see Figure \@ref(fig:pretermcstslacto).

Alternatively, in Figure \@ref(fig:preterm), we avoid working with
CSTs, displaying instead averages among samples associated with either
preterm or term births. The green edges are associated with preterm
births -- we see that they contribute more weight to Phyla outside the
Firmicutes. This is consistent with the claim that CST 4, the most
diverse of the CSTs, is associated with preterm births.

@@

We next consider an application unrelated to the microbiome, but with
relatively clear hierarchical structure. Our data are downloaded from
Zillow^[http://www.zillow.com/research/data/](http://www.zillow.com/research/data/)],
and give the Zillow Home Value Indexes at the neighborhood level,
across the country, given monthly between 1996 and 2016. In our
display, we have taken the base 10 log of these indexes. As our
hierarchical structure, we use each neighborhood's assignment to
state, regional, county, and city levels. We represent each of these
coarser spatial categories using the average of all neighborhoods
contained in them. We have filtered down to the 890 neighborhoods in
California; rendering more neighborhoods while keeping all 246
timepoints causes the interface to lag^[See the Future Work section
for potential optimizations, however.].

Our basic analysis revolves around geographic and temporal variation
in home prices. We are especially interested in the effect of the 2008
recession and any variation in the lead-up or recovery. These
questions can be naturally framed using timebox trees and treeboxes.

We first study variation in prices at the city level. These series can
be highlighted using a single brush in treeboxes, since each spatial
level is displayed at the same height in the tree. THe resulting
display is given in Figure \@ref(fig:zillowcities). From a high level,
the housing bubble, decrease in prices due to the recession, and
subsequent recovery are readily apparent. The city-level view makes it
clear that not all neighborhoods were equally affected by the
recession -- richer cities plateaued at their 2008 prices,
middle-income cities saw moderate decreases, and poorer cities saw the
most significant declines in home prices. These more significant
declines tended to occur in cities that had recently seen rapid
increases during the housing bubble. Finally, we note that the range
in home prices seems to have increased after the recession, speaking
to the differential long-term effect of the recession on prices.

Alternatively, we can consider the trajectories of home prices among
neighborhoods, conditional on their being that were middle-income
before the recession.  We generate the sequence of views in Figures
\@ref(fig:zillowmiddlepre), \@ref(fig:zillowmiddleup), and
\@ref(fig:zillowmiddledown) in this vein. The first of these figures
isolates neighborhoods with middle incomes before the recession, using
a single timebox. Since there appears to be a divergence in
trajectories after the recession, we introduce a second post-recession
timebox, dragging it over series with higher and lower incomes during
this second time period. This is the content of Figures
\@ref(fig:zillowmiddleup) and \@ref(fig:zillowmiddledown). Inspecting
the highlighted tree nodes associated with these series, we find that
most of the middle-income series that increased after the recession
are associated with middle-income neighborhoods within the coastal
Southern California counties. In constrast, those middle income series
that saw decreases were mostly located in Central California and
Oakland.

The previous analysis highlights the fact that, within even narrow geographic
regions, there can be substantial variation in prices. We can study this
directly using treeboxes. In Figure \@ref(fig:zillowsf) we have highlighted all
series in San Francisco County. We see that, in 2016, prices range from around
$10^{5.6} \approx \$400,000$ to $10^{6.7} \approx \$5 \text{million}$. So, while
all these neighborhoods tend to be among the more expensive ones in California,
prices can vary in a non-smooth way across geographic space.

We conclude this example with a caveat that the Zillow data are not
representative of all neighborhoods in California, only those with
enough listings on the site, so should be supplemented by other data
sources for more substantial decision-making.

@@

Our next example is a study in bikesharing demand, included as an
example of analyzing collections of time series when there is no
obvious hierarchical structure a priori. The data are downloaded from the
[UCI Repository](https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset) and
were originally collected by a Washington D.C.-based bikesharing
system for use in a Kaggle prediction competition. The data are hourly
measurements of bike demand, aggregated across all bikesharing
stations, over two years, along with some supplemental weather
data. In the competition, participants were asked to predict the
hourly demand on a held-out test set. Here, we adopt a descriptive
view instead, attempting to characterize factors associated with
variation in bikesharing demand.

Like the Zillow home prices application, we study this problem as one
of describing a large collection of related time series. Here, we
consider the demand during a single day to be one time series; this is
a natural choice considering the daily periodicity of bike demand. To
relate these daily series according to an interpretable tree
structure, we apply a regression tree relating the supplemental data
to the bikesharing demand (J.  Friedman, Hastie, and Tibshirani
2001). In more detail, we build this tree by noting the “two table”
structure of this problem: one describes bike demand, the other holds
the supplemental data. In both, the rows index days, while the columns
index hours and supplemental features for the two tables,
respectively. Our tree is the trained regression tree after predicting
demand at 8AM based on the supplemental data. We choose this response
because (1) we need a univariate response in order to apply regression
trees and (2) the more straightforwards reduction to
daily-average-demand fails to distinguish between weekdays and
weekends, whose series appear qualitatively very different from each
other.

Given this response, the first split in the regression tree is
(unsurprisingly) the difference between weekends and weekdays. This is
emphasized in Figures 14 and 15, respectively; using timeboxes to
isolate the two types of series highlight the left and right sides of
the tree, respectively. For a mroe subtle effect, we select the
internal nodes associated with the first split below the weekday
vs. weekend split; these are given in Figures 16 and 17. This suggests
that more commuters became aware or interested in bikesharing during
the second year of data collection.

To contrast with these general questions about daily demand, we ask a
more granular question about demand at night: what characterizes days
on which there is larger than average demand after midnight? We can
select these series after first zooming into this time window. Figure
18 reveals that the highlighted series are associated with the
warm-weekend split, which seems quite reasonable in retrospect.

Finally, we can study the behavior of the regression tree itself using
the DOI sankey. Here, we group samples according to their quintile of
8AM demand and then count the abundance of the groups flowing down
different branches. We find that the quintiles are each rather
strongly separated after descending even a few steps down the
regression tree – for example, in Figure ?? focuses on 2011 vs. 2012
split among weekday samples, showing that this split distinguishes
between samples falling in the second and third quintiles of 8AM
demand.

This interactive representation of regression trees is potentially
more useful on larger trees that cannot be easily parsed in a single
view; in this sense the bikesharing tree is relatively
simple. Further, in light of research on approximating more complex
supervised learning algorithms using decision trees (Boz 2002, Saito
and Nakano (2002)), this approach of modeling and interpreting
two-table data could be used in conjunction with more sophisticated
regression methods. In our ideas data analysis workflow, we imagine
the analyst applying interactive visualization and modeling techniques
in an iterative, nonlinear fashion.

## Conclusion and Future Work ##

@@

This package has only developed some of the most basic ideas, and
there are a number of potentially useful extensions worth
exploring. For example, we have not considered the principle of
arrangement in our visualizations (Buja, Cook, and Swayne 1996),
though many of our conclusions were based on comparing alternative
selections of the same display. We could imagine faceting our displays
across groups to make these types of comparisons more
accessible. Further, we have only worked with the DOI distribution
described in (Heer and Card 2004). It would be interesting to define a
more statistical notion of interest along nodes, based on cognostics,
for example (Hafen et al. 2013, J. H. Friedman and Stuetzle
(2002)). Further, it could be helpful to link the visualization-based
workflow described in this work with more traditional multivariate and
inferential statistics workflows – it is easy to imagine an
interactive version of ordinations that display representations of the
raw data upon selection. A simple extension could be to allow graph
layouts instead of trees in time and treebox displays, for data that
cannot be coerced into a hierarchical structure. Finally, if these
ideas turn out to be useful in practice, it would be valuable to
modularize the basic visualization layouts and relationships into a
library, allowing the wider community to construct linked, interactive
graphics with minimal effort.

All these potential extensions aside, we have built an easily
accessible R package for visualization techniques in a very specific
methodology problem – analysis of differential abundance and dynamics
in hierarchical structured data – that appears in a variety of
application domains. We have leveraged (Vaidyanathan et al.  2014) to
create exploratory visualizations while still deep in data analysis;
in this way our work is a departure from the culture of polished,
journalistic visualizations prioritized by the D3 community. Finally,
we have given a series of examples to demonstrate how the general
visualization techniques of focus-plus-context and linked brushing can be
practically incorporated into a range of practical analysis workflows,
from studying the impact of microbes on human health to better
allocating units in commuter bikesharing systems.

>>>>

---
title: "Interactive Visualization of Hierarchically Structured Data"
author: "Kris Sankaran"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_book:
    keep_tex: yes
bibliography: ["bibliography.bib"]
link-citations: yes
---

@@

In the DOI sankey, we split each edge in the DOI Tree across different
groups. For example, suppose we have the average counts for treatment
and control groups at each tree node. Every edge in the tree is split
into two colors^[We use the colorbrewer palette to facilitate
readability [@brewer2003colorbrewer]], with relative widths of the
different colors reflecting differences in sizes for the two
groups. The overall width of each edge represents the sum of values
across all groups.

@@

The final display currently implemented in the package is the natural
converse of the timebox trees display. Rather than defining visual
queries in terms of time series, it defines queries using nodes in the
tree. For this reason, we call the display Treeboxes. Rather than
focusing on the intersection of brushes, as in timebox trees, we focus
on the union of brushed over nodes. This allows us to highlight series
associated with nodes on distant subtrees. This display is also suited
for the microbial dynamics problem. For example, by highlighting all
nodes at one taxonomic level in the tree, we can easily summarize the
time series pattern for all the taxa at that level. Alternatively,
focusing on all the children below a single node makes it possible to
see how much correlation and competition there is between
taxonomically similar microbes. As in the timebox trees display, a
search box and pan-zoom scented widget are provided.

@@

We now delve into applications on real data. Our goals are to
illustrate potential workflows that incorporate treelapse, describe
the formulation of questions that can be naturally investigated with
our methods, and provide example interpretations on treelapse
output. Our examples are also chosen to reflect the range of problem
domains to which the package can be applied -- though it was motivated
by applications to the microbiome, it is not tied to it. More
importantly, we argue that the visualization principles reviewed above
can substantively improve the practice of data analysis in the class
of problems to which we have limited ourselves.

@@

[@dethlefsen2008pervasive] investigates the effect of antibiotics on
microbial community composition from an ecological perspective. The
study tracks the microbiome of three patients across ten months, with
two five-day antibiotic time courses separated by 6 months. Discerning
the variation in resilience across microbes is important, considering
the the role of microbes in health and not just disease.

We approach the data using the linked time and treebox views, after
first filtering low variance taxa and taking an $\text{asinh}$
transformation. An initial view, Figure \@ref(fig:antibioticoverview),
reveals two dramatic drops in the overall microbial abundance time
series during the antibiotics time courses. Two more subtle effects
are also suggested from this view,

@@

Next, using the scented widget, we focus on the window around the
second antibiotic treatment. We apply the treebox display to compare
then behavior of different families of Firmicutes, Lachnospiraceae and
Ruminoccocus. We suspect that these taxa are associated with the
delayed recovery after the second time course. To investigate this, we
input these family names in the search box to isolate their positions
on the tree; then we apply brushes to highlight the series that
contribute to these higher-level families. The resulting view is given
by Figure \@ref(fig:antibioticfirmicutes)

@@

Alternatively, we can summarize each node by the average across its
descendants -- this brings attention to individual microbes that may
be underlying some of the broader taxonomic patterns we have noted
when studying the subtree sums. For example, in Figure
\@ref(fig:ruminococcus), we highlight all families below order
Ruminococcus, suggesting that the decrease due to antibiotics occurs
uniformly across almost all families. A point that was not evident in
the earlier sum-across-descendants view is that, after the second
treatment of antibiotics, a few of the Ruminoccocus families recover
more rapidly than the rest, for example the Unc095d3 (highlighted in
red) are only briefly affected. In contrast, most families seem to
recover in unison after the first treatment.

@@

Further, note that in this subtree averages view, the tree display has
changed. This is because, at each branching point, we place the node
with larger average value on the left. This places the
Verrucomicrobiae at the far left, which seem to have large abundance
over many time points. This phylum had been previously obfuscated --
because there are not many leaves associated with this phylum, the sum
was small. Interestingly, the abundance of these microbes seems to
*increase* after the first antibiotics treatment.  Be cautious,
however, that the average over only a few Verrucomicrobiae species
will be a more variable estimate than the averages over the more
prevalent phyla.

@@

The study [@digiulio2015temporal] tracked the abundance of microbes in
the vaginal microbiome during pregnancy in an effort to study
relationships between microbial community composition and preterm
birth. Ideally, it would be possible to develop clear microbial
signatures associated with preterm births.

Unlike the antibiotics study, we have measurements across more
individuals than we could reasonably inspect one at a time. While we
could average across all individuals, we will take our cue from
[@digiulio2015temporal] and place each sample into one of five
Community State Types (CSTs), identified via k-medoids.  In that
study, a linear model identified one of these CSTs (CST 4) as
significantly more diverse, further it appeared associated with
preterm births. Here, we corroborate this finding using exploratory
views.

Therefore, our focus here is on the differential abundance question,
rather than dynamics.  We would like to provide visual representations
of differential abundance across CSTs and also between preterm and
non-preterm births. [@digiulio2015temporal] interpreted the CSTs using
a heatmap, with microbes ordered according to a hierarchical
clustering.  By using the DOI sankey instead, we can interpret the
CSTs in their taxonomic context and at multiple scales of taxonomic
resolution. Further, while [@digiulio2015temporal] focused on
identifying associations between preterm births and CSTs -- presumably
because testing individual microbes loses power -- we can compare
microbe abundances between preterm and non-preterm samples along
subtrees, without requiring CSTs as an intermediary.

In Figure \@ref(fig:pretermcsts), we compare the 5 CSTs according to
their values along the subtree. Specifically, we took the average of
all samples within each CST to define values at the (microbe-level)
leaf nodes, and then aggregated the averages up to the root. It is
immediately clear that samples from CST 4 have much more taxonomic
diversity. Further, focusing on the Lactobacillaceae family, we note
that the differential abundance of these microbes distinguishes the
remaining CSTs, see Figure \@ref(fig:pretermcstslacto).

Alternatively, in Figure \@ref(fig:pretermcsts), we avoid working with
CSTs, displaying instead averages among samples associated with either
preterm or term births. The green edges are associated with preterm
births -- we see that they contribute more weight to phyla outside the
Firmicutes. This is consistent with the claim that CST 4, the most
diverse of the CSTs, is associated with preterm births.

@@

We next consider an application unrelated to the microbiome, but with
relatively clear hierarchical structure. Our data are downloaded from
Zillow[^[http://www.zillow.com/research/data/](http://www.zillow.com/research/data/)]],
and give the Zillow Home Value Indexes at the neighborhood level,
across the country, computed monthly between 1996 and 2016. In our
display, we have taken the base 10 log of these indexes. As our
hierarchical structure, we use each neighborhood's assignment to
state, regional, county, and city levels. We represent each of these
coarser spatial categories using the average of all neighborhoods
contained in them. We have filtered down to the 890 neighborhoods in
California; rendering more neighborhoods while keeping all 246
timepoints causes the interface to lag^[See the Section
\@ref(conclusion) for potential optimizations, however.].

Our basic analysis revolves around geographic and temporal variation
in home prices. We are especially interested in the effect of the 2008
recession and any variation in the lead-up to or recovery from this
event. These questions can be naturally framed using timebox trees and
treeboxes.

We first study variation in prices at the city level. These series can
be highlighted using a single brush in treeboxes, since each spatial
level is displayed at the same height in the tree. The resulting
display is given in Figure \@ref(fig:zillowcities). From a high level,
the housing bubble, decrease in prices due to the recession, and
subsequent recovery are readily apparent. The city-level view makes it
clear that not all neighborhoods were equally affected by the
recession -- richer cities plateaued at their 2008 prices,
middle-income cities saw moderate decreases, and poorer cities saw the
most significant declines in home prices. These more significant
declines tended to occur in cities that had recently seen rapid
increases during the housing bubble. Finally, we note that the range
in home prices seems to have increased after the recession, speaking
to the differential long-term effect of the recession on prices.

Alternatively, we can study the trajectories of home prices among
neighborhoods, conditional on their being middle-income before the
recession. We generate the sequence of views in Figures
\@ref(fig:zillowmiddlepre), \@ref(fig:zillowmiddleup), and
\@ref(fig:zillowmiddledown) to this end. The first of these figures
isolates neighborhoods with middle incomes before the recession, using
a single timebox. Since there appears to be a divergence in
trajectories after the recession, we introduce a second post-recession
timebox, dragging it over series with higher and lower incomes during
this second time period. This is the content of Figures
\@ref(fig:zillowmiddleup) and \@ref(fig:zillowmiddledown). Inspecting
the highlighted tree nodes associated with these series, we find that
most of the middle-income series that increased after the recession
are associated with middle-income neighborhoods within the coastal
Southern California counties. In constrast, those middle income series
that saw decreases were mostly located in Central California and
Oakland.

The previous analysis highlights the fact that, within even narrow geographic
regions, there can be substantial variation in prices. We can study this
directly using treeboxes. In Figure \@ref(fig:zillowsf) we have highlighted all
series in San Francisco County. We see that, in 2016, prices range from around
$10^{5.6} \approx \$400,000$ to $10^{6.7} \approx \$5 \text{ million}$. So, while
all these neighborhoods tend to be among the more expensive ones in California,
prices can vary in a non-smooth way across geographic space.

We conclude this example with a caveat that the Zillow data are not
representative of all neighborhoods in California, only those with
enough listings on the site, so should be supplemented by other data
sources for any substantial decision-making.

@@

Our next example is a study in bikesharing demand, included as an
example of analyzing collections of time series when there is no
obvious hierarchical structure a priori. The data are available at
the [UCI
Repository][^[(https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset)](https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset)]
and were originally collected by a Washington D.C.-based bikesharing
system for use in a Kaggle prediction competition. The data are hourly
measurements of bike demand, aggregated across all bikesharing
stations, over two years, along with supplemental weather data. In the
competition, participants were asked to predict the hourly demand on a
held-out test set. Here, we adopt a descriptive view instead,
attempting to characterize factors associated with variation in
bikesharing demand.

Like the Zillow home prices application, we study this problem as one
of describing a large collection of related time series. Here, we
consider the demand during a single day to be one time series; this is
a natural choice considering the daily periodicity of bike demand. To
arrange these daily series along an interpretable tree structure, we
apply a regression tree relating the supplemental data to the
bikesharing demand [@breiman1984classification]. In more detail, we
built this tree by noting the “two table” structure of this problem:
one describes bike demand, the other holds the supplemental data. In
both, the rows index days, while the columns index either hours or
supplemental features. Our tree is the trained regression tree after
predicting demand at 8AM based on the supplemental data. We choose
this response because (1) we need a univariate response in order to
apply regression trees and (2) the more straightforwards reduction to
daily-average-demand fails to distinguish between weekdays and
weekends, whose series appear qualitatively very different from each
other.

Given this response, the first split in the regression tree is
(unsurprisingly) the difference between weekends and weekdays. This is
emphasized in Figures 14 and 15, respectively; using timeboxes to
isolate the two types of series highlight the left and right sides of
the tree, respectively. For a more subtle effect, we select the
internal nodes associated with the first split below the weekday
vs. weekend split; these are given in Figures 16 and 17. This suggests
that weekday demand increased during the second year.

In contrast to these general questions about daily demand, we could
ask a more granular question about specific time windows. For example,
what characterizes days on which there is larger than average demand
after midnight? We can select these series after first zooming into
this time window. Figure 18 reveals that the highlighted series are
associated with the warm-weekend split, which seems quite reasonable
in retrospect.

Finally, we can study the behavior of the regression tree itself using
the DOI sankey. Here, we group samples according to their quintile of
8AM demand and then count the abundance of the groups flowing down
different branches. We find that the quintiles are each rather
strongly separated after descending even a few steps down the
regression tree – for example, Figures \@ref(fig:weekday_2011) and
\@ref(fig:weekday_2012) focuse on 2011 vs. 2012 split among weekday
samples, showing that this split distinguishes between samples falling
in the second and third quintiles of 8AM demand.

This interactive representation of regression trees is potentially
more useful on larger trees that cannot be easily parsed in a single
view; in this sense the bikesharing tree is relatively
simple. Further, in light of research on approximating more complex
supervised learning algorithms using decision trees (Boz 2002, Saito
and Nakano (2002)), this approach of modeling and interpreting
two-table data could be used in conjunction with more sophisticated
regression methods. In our ideal data analysis workflow, we imagine
the analyst applying interactive visualization and modeling techniques
in an iterative, nonlinear fashion.

```{r working, fig.cap = "The two peaks at rush hour distinguish weekday series from the rest."}
knitr::include_graphics("working.png")
```

```{r weekend, fig.cap = "Unlike weekday demand, weekend demand is unimodal. The few weekday series with unimodal series seem to be associated with holidays. This is the case for New Years' Eve, which is currently hovered over in the tree."}
knitr::include_graphics("weekend.png")
```

```{r weekday_2011, fig.cap = "Weeday demand appears larger in 2012 than 2011 -- compare with the next figure."}
knitr::include_graphics("weekday_2011.png")
```

```{r weekday_2012, fig.cap = "Weekday demand increased in 2012 -- compare with the previous figure."}
knitr::include_graphics("weekday_2012.png")
```

```{r warm_weekend, fig.cap = "The samples with highest night demand tend to fall on warm weekends."}
knitr::include_graphics("warm_weekend.png")
```

```{r bike_sankey, fig.cap = "We can interpret the regression tree using an interactive DOI representation."}
knitr::include_graphics("bike_sankey.png")
```

## Conclusion and Future Work {#conclusion}

@@

This package has only developed basic ideas, and there are a number of
potentially useful extensions worth exploring. For example, we have
not considered the principle of arrangement in our visualizations
(Buja, Cook, and Swayne 1996), though many of our conclusions were
based on comparing alternative selections of the same display. We
could imagine faceting our displays across groups to make these types
of comparisons more accessible. Further, we have only worked with the
DOI distribution described in (Heer and Card 2004). It would be
interesting to define a more statistical notion of interest along
nodes, based on cognostics, for example (Hafen et al. 2013,
J. H. Friedman and Stuetzle (2002)). A simple extension could be to
allow graph layouts instead of trees in time and treebox displays, for
data that cannot be coerced into a hierarchical structure. Finally, if
these ideas turn out to be useful in practice, it would be valuable to
modularize the basic visualization layouts and relationships into a
library, allowing the wider community to construct novel linked,
interactive graphics with minimal effort.

In summary, we have built an easily accessible R package for
visualization techniques in a very specific methodology problem –
analysis of differential abundance and dynamics in hierarchically
structured data – that appears in a variety of application domains. We
have leveraged (Vaidyanathan et al.  2014) to create exploratory
visualizations while still deep in data analysis; in this way our work
is a departure from the culture of polished, journalistic
visualizations prioritized by the D3 community. Finally, we have given
a series of examples to demonstrate how the general visualization
techniques of focus-plus-context and linked brushing can be
practically incorporated into a range of practical analysis workflows,
from studying the impact of microbes on human health to better
allocating units in commuter bikesharing systems.
